---
title: "Wildlife Rehabilitation: analyses used in Davy et al."
author: "James E Paterson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

This document contains the analyses for Davy et al. modelling the effects of wildlife rehabilitation on population trajectory and extinction risk.

We used five model species:

* Blanding's Turtles
* Snapping Turtles
* Painted Turtles
* Common Raccoons
* Little Brown Myotis

This document is broken down by sections in the Methods & Results

1. Effects of additive mortality
2. Effects of rehabilitation

```{r loadpackages, echo = FALSE, message = FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(png)
library(grid)
```

## 1. Effects of additive adult mortality

We used Population Viability Analyses (PVAs) in the program VORTEX to test how increases in mortality affect populations and extinction risk (this section) and whether that risk can be mitigated by rehabilitation (next section).

First, we loaded all the output data from the Vortex simulations.

```{r loadvortexdata, messsage = FALSE, warning = FALSE}
# Load data (csv file with cleaned output summaries for all scenarios)
wr.data <- read.csv(file = "Data/wrvortexdata_clean.csv")#,
                    #stringsAsFactors = FALSE)

# Ordering scenario name (important for colours and legends to properly line-up)
wr.data$scen.name2 <- factor(wr.data$scen.name2, levels = c("base", "+1% harvest",
                                            "+2% harvest", "+5% harvest",
                                             "+1% harvest +10% rehab","+1% harvest +25% rehab","+1% harvest +50% rehab",
                                            "+2% harvest +10% rehab","+2% harvest +25% rehab", "+2% harvest +50% rehab",
                                            "+5% harvest +10% rehab", "+5% harvest +25% rehab", "+5% harvest +50% rehab"))
```

We compared the population trajectory for all species under 4 scenarios:

1. Base (using long-term demographic estimates; see Table S1 in supplementary information for values and sources)
2. An increase in adult mortality of 1%
3. An increase in adult mortality of 2%, and
4. An increase in adult mortality of 5%

```{r mortalityeffects, echo = FALSE, message = FALSE}
# Titles for plots
racc.label <- expression(paste("A) Raccoon ", italic("Procyon lotor"), sep = " "))
patu.label <- expression(paste("B) Painted Turtle ", italic("Chrysemys picta"), sep = " "))
sntu.label <- expression(paste("C) Snapping Turtle ", italic("Chelydra serpentina"), sep = " "))
bltu.label <- expression(paste("D) Blanding's Turtle ", italic("Emydoidea blandingii"), sep = " "))
lbba.label <- expression(paste("E) Little Brown Myotis ", italic("Myotis lucifugus"), sep = " "))

# Load outlines for plots
patu.ima <- readPNG("Figures/outlines/patu_outline.png")
patu.g <- rasterGrob(patu.ima, interpolate = TRUE)

bltu.ima <- readPNG("Figures/outlines/leifso_bltu_silhouette.png")
bltu.g <- rasterGrob(bltu.ima, interpolate = TRUE)

sntu.ima <- readPNG("Figures/outlines/sntu_outline_dark.png")
sntu.g <- rasterGrob(sntu.ima, interpolate = TRUE)

racc.ima <- readPNG("Figures/outlines/racc_outline.png")
racc.g <- rasterGrob(racc.ima, interpolate = TRUE)

lbba.ima <- readPNG("Figures/outlines/myotis_outline.png")
lbba.g <- rasterGrob(lbba.ima, interpolate = TRUE)

# Painted Turtle plot: additive mortality
figure1.patu.plot <- ggplot(wr.data[wr.data$species == "PATU" & wr.data$rehab == "none",], 
                       aes(Year, Nall)) +
  ggtitle(patu.label) +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_vline(aes(xintercept = 3*11.59), linetype = "dashed", col = "black") +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("green","#0072B2", "#000000","#D55E00" ),
                      labels = c("None", "+1%", "+2%", "+5%"),
                      name = "Change in\nadult mortality") +
  # annotation_custom(gpp, xmin = 0, xmax = 20, ymax = 400, ymin = 325) +
  theme_classic() +
  annotation_custom(patu.g, xmin = -10, xmax = 40, ymin = 800, ymax = 1000) +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1000))
# figure1.patu.plot


# Now for snapping turtles
figure1.sntu.plot <- ggplot(wr.data[wr.data$species == "SNTU" & wr.data$rehab == "none",], 
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*26.06), linetype = "dashed", col = "black") +
  ggtitle(sntu.label) +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("green","#0072B2", "#000000","#D55E00" ),
                      labels = c("None", "+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  annotation_custom(sntu.g, xmin = -5, xmax = 55, ymin = 800, ymax = 1000) +
  theme_classic() +
  theme(legend.position = "none") + # c(0.30, 0.77)
  coord_cartesian(ylim = c(0, 1000))
# figure1.sntu.plot

# Blanding's Turtle plot: additive mortality
figure1.bltu.plot <- ggplot(wr.data[wr.data$species == "BLTU" & wr.data$rehab == "none",], 
                       aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*35.11), linetype = "dashed", col = "black") +
  ggtitle(bltu.label) +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("green","#0072B2", "#000000","#D55E00" ),
                      labels = c("None", "+1%", "+2%", "+5%"),
                      name = "Change in\nadult mortality") +
  theme_classic() +
  annotation_custom(bltu.g, xmin = -5, xmax = 55, ymin = 800, ymax = 1000) +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1000))
# figure1.bltu.plot

figure1.racc.plot <- ggplot(wr.data[wr.data$species == "Racc" & wr.data$rehab == "none" ,], 
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*2.98), linetype = "dashed", col = "black") +
  ggtitle(racc.label) +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("green","#0072B2", "#000000","#D55E00" ),
                      labels = c("None", "+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  annotation_custom(racc.g, xmin = 5, xmax = 60, ymin = 875, ymax = 1050) +
  theme_classic() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1000))
# figure1.racc.plot

# LBBA
figure1.lbba.plot <- ggplot(wr.data[wr.data$species == "LBBA" & wr.data$rehab == "none",], 
                             aes(Year, Nall)) +
  ggtitle(lbba.label) +
  geom_vline(aes(xintercept = 3*6.67), linetype = "dashed", col = "black") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("green","#0072B2", "#000000","#D55E00" ),
                      labels = c("None", "+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  annotation_custom(lbba.g, xmin = 5, xmax = 65, ymin = 800, ymax = 1000) +
  theme_classic() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1000))
# figure1.lbba.plot

legend <-lemon::g_legend(figure1.patu.plot + theme(legend.position='right',
                                                    legend.text=element_text(size=20),
                                                    legend.title=element_text(size=20)))


gridExtra::grid.arrange(figure1.racc.plot, figure1.bltu.plot, 
                        figure1.patu.plot, figure1.lbba.plot,
                        figure1.sntu.plot, legend,
                        ncol = 2, nrow = 3)



# Save
# 1. Open file
png("Figures/Figure1_trajectory_adultmortality.png", width = 9, height = 11, units = "in", res = 500)
# 2. Create plot
gridExtra::grid.arrange(figure1.racc.plot, figure1.bltu.plot, 
                        figure1.patu.plot, figure1.lbba.plot,
                        figure1.sntu.plot, legend,# the legend grob here
                        ncol = 2, nrow = 3)
# 3. Close the file
dev.off()
```

It is clear that populations of long-lived species (turtles and bats) are sensitive to small increases in adult mortality. For turtles, these mortality rates are expected based on traffic volume, road density, and wetland distribution in much of eastern North America. An increase of 5% is unsustainable for all three turtle species. For Blanding's Turtles and Snapping Turtles (which mature later than Painted Turtles and have lower juvenile survival), even increases in adult mortality of 2% lead to population declines. Little Brown Bats declined when adult mortality was 5% per year.

```{r createextinctiontable, echo = FALSE, message = FALSE, warning = FALSE}
# Make a table with extinction probabilities for each scenario
pext.df <- wr.data[wr.data$Year == 200,]
pext.df$rehab <- factor(pext.df$rehab, levels = c("none","+10%", "+25%", "+50%"))

pext.df %>%
  filter(rehab == "none") %>%
  select(species, PExtant, group) %>%
  mutate(PercentExtinct = (1 - PExtant)*100) %>%
  arrange(species, PercentExtinct) %>%
  knitr::kable()
```

## 2. Effects of wildlife rehabilitation

To test how wildlife rehabilitation affects wildlife populations, we modelled 3 different options for each mortality scenario:

1. 10% of injured adults are rehabilitated
2. 25% of injured adults are rehabilitated
3. 50% of injured adults are rehabilitated.

See the next section for a validation of these proportions, based on intake data of injured turtles to the Ontario Turtle Conservation Centre.

Below, we plot how wildlife rehabilitation affects each species.

```{r raccrehabgraph, echo = FALSE, message = FALSE}
# Figure 2. 3 plots per species showing effects of rehabilitation on simulations with 1, 2, and 5% harvest rate
racc.1.plot <- ggplot(wr.data[wr.data$species == "Racc" & 
                                       wr.data$group == "Racc +1% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*2.98), linetype = "dashed", col = "black") +
  ggtitle("A) +1% mortality") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = rev(c("#D55E00", "#000000", "#0072B2","green")),
                      labels = rev(c("0% rehabbed", "+10% rehabbed", 
                                 "+25% rehabbed", "+50% rehabbed")),
                      name = "Scenario") +
  annotation_custom(racc.g, xmin = 15, xmax = 90, ymax = 1050, ymin = 875) +
  theme_classic() +
  theme(legend.position =  c(0.65, 0.325),
        legend.title = element_blank()) +
  coord_cartesian(ylim = c(0, 1000))
# racc.1.plot

racc.2.plot <- ggplot(wr.data[wr.data$species == "Racc" & 
                                       wr.data$group == "Racc +2% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*2.98), linetype = "dashed", col = "black") +
  ggtitle("B) +2% mortality") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+2% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position =  "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# racc.2.plot

racc.5.plot <- ggplot(wr.data[wr.data$species == "Racc" & 
                                       wr.data$group == "Racc +5% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*2.98), linetype = "dashed", col = "black") +
  ggtitle("C) +5% mortality") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+5% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# racc.5.plot
```


```{r paturehabgraph, echo = FALSE, message = FALSE, warning = FALSE}
patu.1.plot <- ggplot(wr.data[wr.data$species == "PATU" & 
                                       wr.data$group == "PATU +1% harvest",],
                           aes(Year, Nall)) +
  ggtitle("D)") +
  geom_vline(aes(xintercept = 3*11.59), linetype = "dashed", col = "black") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+1% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "Scenario") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000)) +
  annotation_custom(patu.g, xmin = 0, xmax = 75, ymin = 800, ymax = 1000)
# patu.1.plot

patu.2.plot <- ggplot(wr.data[wr.data$species == "PATU" & 
                                       wr.data$group == "PATU +2% harvest",],
                             aes(Year, Nall)) +
  ggtitle("E)") +
  geom_vline(aes(xintercept = 3*11.59), linetype = "dashed", col = "black") +
   geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+2% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# patu.2.plot

patu.5.plot <- ggplot(wr.data[wr.data$species == "PATU" & 
                                       (wr.data$group == "PATU +5% harvest" | wr.data$group == "PATU +5%harvest"),],
                             aes(Year, Nall)) +
  ggtitle("F)") +
  geom_vline(aes(xintercept = 3*11.59), linetype = "dashed", col = "black") +
   geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+5% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# patu.5.plot
```


```{r blturehabgraph, echo = FALSE, message = FALSE, warning = FALSE}
bltu.1.plot <- ggplot(wr.data[wr.data$species == "BLTU" & 
                                       wr.data$group == "BLTU +1% harvest",],
                           aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*35.11), linetype = "dashed", col = "black") +
  ggtitle("J)") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+1% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "Scenario") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000)) +
  annotation_custom(bltu.g, xmin = -10, xmax = 95, ymin = 750, ymax = 1000)
# bltu.1.plot

bltu.2.plot <- ggplot(wr.data[wr.data$species == "BLTU" & 
                                       wr.data$group == "BLTU +2% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*35.11), linetype = "dashed", col = "black") +
  ggtitle("K)") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+2% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# bltu.2.plot

bltu.5.plot <- ggplot(wr.data[wr.data$species == "BLTU" & 
                                       (wr.data$group == "BLTU +5% harvest" | wr.data$group == "BLTU +5%harvest"),],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*35.11), linetype = "dashed", col = "black") +
  ggtitle("L)") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00",  "#000000","#0072B2", "green"),
                      labels = c("+5% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# bltu.5.plot
```

```{r snturehabgraph, echo = FALSE, message = FALSE}
sntu.1.plot <- ggplot(wr.data[wr.data$species == "SNTU" & 
                                       wr.data$group == "SNTU +1% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*26.06), linetype = "dashed", col = "black") +
  ggtitle("G)") +
   geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+1% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "Scenario") +
  annotation_custom(sntu.g, xmin = -5, xmax = 110, ymax = 1000, ymin = 700) +
  theme_classic() +
  theme(legend.position =  "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# sntu.1.plot

sntu.2.plot <- ggplot(wr.data[wr.data$species == "SNTU" & 
                                       wr.data$group == "SNTU +2% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*26.06), linetype = "dashed", col = "black") +
  ggtitle("H)") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+2% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position =  "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# sntu.2.plot

sntu.5.plot <- ggplot(wr.data[wr.data$species == "SNTU" & 
                                       wr.data$group == "SNTU +5% harvest",],
                             aes(Year, Nall)) +
  geom_vline(aes(xintercept = 3*26.06), linetype = "dashed", col = "black") +
  ggtitle("I)") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+5% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# sntu.5.plot
```

```{r lbbrehabgraph, echo = FALSE, message = FALSE}
lbba.1.plot <- ggplot(wr.data[wr.data$species == "LBBA" & 
                                       wr.data$group == "LBBA +1% harvest",],
                             aes(Year, Nall)) +
  ggtitle("M)") +
  geom_vline(aes(xintercept = 3*6.67), linetype = "dashed", col = "black") +
   geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = rehab), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      name = "Scenario") +
  annotation_custom(lbba.g, xmin = 0, xmax = 75, ymax = 1000, ymin = 800) +
  theme_classic() +
  theme(legend.position =  "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# lbba.1.plot

lbba.2.plot <- ggplot(wr.data[wr.data$species == "LBBA" & 
                                       wr.data$group == "LBBA +2% harvest",],
                             aes(Year, Nall)) +
  ggtitle("N)") +
  geom_vline(aes(xintercept = 3*6.67), linetype = "dashed", col = "black") +
  geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+2% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position =  "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# lbba.2.plot

lbba.5.plot <- ggplot(wr.data[wr.data$species == "LBBA" & 
                                       wr.data$group == "LBBA +5% harvest",],
                             aes(Year, Nall)) +
  ggtitle("O)") +
  geom_vline(aes(xintercept = 3*6.67), linetype = "dashed", col = "black") +
   geom_ribbon(aes(ymin = Nall-1.96*SE.Nall., ymax = Nall+1.96*SE.Nall., x = Year, group = scen.name), 
              alpha = 0.3) +
  geom_line(aes(colour = scen.name2)) +
  labs(y = "Population size", colour = "Scenario") +
  scale_colour_manual(values = c("#D55E00", "#000000", "#0072B2","green"),
                      labels = c("+5% mortality", "+10% adults rehabbed", 
                                 "+25% adults rehabbed", "+50% adults rehabbed"),
                      name = "") +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1000))
# lbba.5.plot
```


```{r displayrehabplots, echo = FALSE, message = FALSE, warning = FALSE}
# Display all plots in a grid for Figure 2
gridExtra::grid.arrange(racc.1.plot, patu.1.plot, sntu.1.plot, bltu.1.plot, lbba.1.plot,
                        racc.2.plot, patu.2.plot, sntu.2.plot, bltu.2.plot, lbba.2.plot,
                        racc.5.plot, patu.5.plot, sntu.5.plot, bltu.5.plot, lbba.5.plot,
                        ncol = 5)

# Save
# 1. Open file
png("Figures/Figure2_rehab_trajectory.png", width = 11, height = 8, units = "in", res = 500)
# 2. Create plot
gridExtra::grid.arrange(racc.1.plot, patu.1.plot, sntu.1.plot, bltu.1.plot, lbba.1.plot,
                        racc.2.plot, patu.2.plot, sntu.2.plot, bltu.2.plot, lbba.2.plot,
                        racc.5.plot, patu.5.plot, sntu.5.plot, bltu.5.plot, lbba.5.plot,
                        ncol = 5)
# 3. Close the file
dev.off()
```

These plots clearly show that:

1. Populations are sensistive to increased mortality rates, and 
2. Rehabilitating adults (and thus reducing mortality rates) can slow down or mitigate declines

How does wildlife rehabilitation affect the probability of extinction? In other words, how do different levels of wildlife rehabilitation affect the proportion of iterations that go extinct for each scenario?

```{r calculateplotextinctionrisk, echo = FALSE, message = FALSE}
patu.ext <- ggplot(pext.df[pext.df$species == "PATU" & pext.df$group != "PATU base",],
                   aes(x = rehab, y = PExtant, col = group)) +
  geom_point() +
  ggtitle(patu.label) +
  geom_line(aes(group = group)) +
  geom_errorbar(aes(ymax = PExtant+SE.PExtant., ymin = PExtant-SE.PExtant.),
                width = 0) +
  scale_colour_manual(values = c("#0072B2","#000000", "#D55E00"),
                      labels = c("+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  labs(x = "Rehabilitation scenario", y = "Probability of persistence (within 200 years)",
       colour = "Amount of additive mortality") +
  annotation_custom(patu.g, xmin = 3.75, xmax = 4.5, ymax = 0.20, ymin = 0) +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1.05))
# patu.ext

bltu.ext <- ggplot(pext.df[pext.df$species == "BLTU" & pext.df$group != "BLTU base",],
                       aes(x = rehab, y = PExtant, col = group)) +
  geom_point() +
  ggtitle(bltu.label) +
  geom_line(aes(group = group)) +
  geom_errorbar(aes(ymax = PExtant+SE.PExtant., ymin = PExtant-SE.PExtant.),
                  width = 0) +
  scale_colour_manual(values = c("#0072B2","#000000", "#D55E00"),
                      labels = c("+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  labs(x = "", y = "Probability of persistence (within 200 years)") +
  theme_classic() +
  annotation_custom(bltu.g, xmin = 3.0, xmax = 4.5, ymax = 0.35, ymin = -0.2) +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1.05))
# bltu.ext

sntu.ext <- ggplot(pext.df[pext.df$species == "SNTU" & pext.df$group != "SNTU base",],
                   aes(x = rehab, y = PExtant, col = group)) +
  geom_point() +
  ggtitle(sntu.label) +
  geom_line(aes(group = group)) +
  geom_errorbar(aes(ymax = PExtant+SE.PExtant., ymin = PExtant-SE.PExtant.),
                width = 0) +
  scale_colour_manual(values = c("#0072B2","#000000", "#D55E00"),
                      labels = c("+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  labs(x = "Rehabilitation scenario", y = "Probability of persistence (within 200 years)",
       colour = "Amount of additive mortality") +
  annotation_custom(sntu.g, xmin = 3.0, xmax = 4.5, ymax = 0.35, ymin = -0.1) +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1.05))
# sntu.ext

lbba.ext <- ggplot(pext.df[pext.df$species == "LBBA" & pext.df$group != "LBBA base",],
                   aes(x = rehab, y = PExtant, col = group)) +
  geom_point() +
  ggtitle(lbba.label) +
  geom_line(aes(group = group)) +
  geom_errorbar(aes(ymax = PExtant+SE.PExtant., ymin = PExtant-SE.PExtant.),
                width = 0) +
  scale_colour_manual(values = c("#0072B2","#000000", "#D55E00"),
                      labels = c("+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  labs(x = "Rehabilitation scenario", y = "Probability of persistence (within 200 years)",
       colour = "Amount of additive mortality") +
  annotation_custom(lbba.g, xmin = 3.5, xmax = 4.5, ymax = 0.25, ymin = 0) +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1.05))
# lbba.ext

racc.ext <- ggplot(pext.df[pext.df$species == "Racc" & pext.df$group != "Racc base",],
                   aes(x = rehab, y = PExtant, col = group)) +
  geom_point() +
  ggtitle(racc.label) +
  geom_line(aes(group = group)) +
  geom_errorbar(aes(ymax = PExtant+SE.PExtant., ymin = PExtant-SE.PExtant.),
                width = 0) +
  scale_colour_manual(values = c("#0072B2","#000000", "#D55E00"),
                      labels = c("+1%", "+2%", "+5%"),
                      name = "Change in adult mortality") +
  labs(x = "Rehabilitation scenario", y = "Probability of persistence (within 200 years)",
       colour = "Amount of additive mortality") +
  annotation_custom(racc.g, xmin = 3.5, xmax = 4.5, ymax = 0.25, ymin = 0) +
  theme_classic() +
  theme(legend.position = "hidden") +
  coord_cartesian(ylim = c(0, 1.05))
# racc.ext

# Make legend
ext.legend <- lemon::g_legend(racc.ext + 
                                theme(legend.position = 'right', 
                                      legend.text = element_text(size = 20),
                                      legend.title = element_text(size = 20)))


gridExtra::grid.arrange(racc.ext, bltu.ext, 
                        patu.ext, lbba.ext,
                        sntu.ext, ext.legend,
                        ncol = 2, nrow = 3)

# Save
# 1. Open file
png("Figures/Figure3_extinctionrisks.png", width = 9, height = 11, units = "in", res = 500)
# 2. Create plot
gridExtra::grid.arrange(racc.ext, bltu.ext, 
                        patu.ext, lbba.ext,
                        sntu.ext, ext.legend,
                        ncol = 2, nrow = 3)
# 3. Close the file
dev.off()
```

For both species, there was almost no risk of extinction when mortality was increased by 1% (regardless of rehabilitation levels).

```{r extinctiontable1, echo = FALSE, message = FALSE, warning = FALSE}
mortality1.table <- pext.df %>%
  filter(group %in% c("BLTU +1% harvest", "SNTU +1% harvest")) %>%
  select(species, rehab, PExtant) %>%
  mutate(PercentExtinct= 100*(1-PExtant))

knitr::kable(mortality1.table)
```

However, at 2% increases in mortality, rehabilitation reduced the chances of extinction within 200 years, especially for snapping turtles (from 16% with no rehab to <2% for rehabbing 50% of struck turtles).

```{r extinctiontable2, echo = FALSE, message = FALSE, warning = FALSE}
mortality2.table <- pext.df %>%
  filter(group %in% c("BLTU +2% harvest", "SNTU +2% harvest")) %>%
  select(species, rehab, PExtant) %>%
  mutate(PercentExtinct= 100*(1-PExtant))

knitr::kable(mortality2.table)
```

At increases of mortality of 5%, the probability of extinction within 200 years was high for both Blanding's Turtles (92%) and Snapping Turtles (88%). Even small amounts of rehabilitation (10% of struck adults) led to decreases in the risk of extinction within 200 years (82% in Blanding's Turtles, 80% in Snapping Turtles).

```{r extinctiontable5, echo = FALSE, message = FALSE, warning = FALSE}
mortality5.table <- pext.df %>%
  filter(group %in% c("BLTU +5% harvest", "SNTU +5% harvest")) %>%
  select(species, rehab, PExtant) %>%
   mutate(PercentExtinct= 100*(1-PExtant))

knitr::kable(mortality5.table)
```

Another useful way to examine the results of the probabilities of persistence is to use a heat map.

```{r persistenceheatmap, message = FALSE, comment = FALSE, echo = FALSE, warning = FALSE}
# First need some re-shaping
library(tidyr)
library(kableExtra)

persistence.table <- pext.df %>%
  mutate(newgroup = paste(species,scen.name2, sep = " "))  %>%
  filter(!(scen.name2 %in% c("+20% harvest", "+30% harvest"))) %>%
  select(species, scen.name2, PExtant) %>%
  spread(species, PExtant) %>%
  mutate(group = substr(scen.name2, 1, 4),
         rehab = substr(scen.name2, 13, 16)) %>%
  select(group, rehab, Racc, PATU, BLTU, SNTU, LBBA) %>%
  arrange(group, rehab)

persistence.table$rehab[persistence.table$rehab == ""] <- "none"

persistence.table$rehab <- factor(persistence.table$rehab,
                                  levels = c("+50%", "+25%","+10%", "none"))

persistence.table$group <- factor(persistence.table$group, 
                                     levels =  c("base", "+1% ", "+2% ", "+5% "))

persistence.table <- persistence.table %>% 
  arrange(group, rehab)

# Change names
names(persistence.table) <- c("Adult mortality", "Rehabilitation", 
                              "Raccoon", "Painted Turtle", "Blanding's Turtle",
                              "Snapping Turtle", "Little Brown Bat")

# # Version 1
# pal.fnc = colorRamp(c("red", "yellow"))
# 
# max.val <- max(persistence.table[ , sapply(persistence.table, is.numeric)], na.rm=TRUE)
# 
# 
# persistence.table2 <- persistence.table %>%
#   # cell_spec(color = "black") %>%
#     mutate_if(is.numeric, function(x) {
#   cell_spec(round(x,2), bold = T, color = "black",
#             background = rgb(pal.fnc(x/max.val) %>% replace(., is.na(.), 200), maxColorValue=255))
# })
# 
# kable(persistence.table2,
#       escape = F ,
#       align = "c"
#       ) %>%
#   kable_styling(c("condensed"), full_width = F)
# 
# # Version 2
# persistence.table %>%
#   mutate_if(is.numeric, function(x) {
#     cell_spec(x*100, bold = T,
#               background = spec_color(x*100, begin = 0, end = 1,
#                                       option = "D", direction = 1),
#               color = "white")
#   }) %>%
#   kable(escape = F, align = "c") %>%
#   kable_styling(c("striped", "condensed"), full_width = F)


# ggplot version
persistence.table3 <- reshape2::melt(persistence.table)
persistence.table3$group <- paste(persistence.table3$'Adult mortality',
                                  persistence.table3$Rehabilitation,
                                  sep = " ")


rehab.labels.df <- data.frame(rehab.labels = rev(c("none",
               "10%",
               "25%",
               "50%",
               "none",
               "10%",
               "25%",
               "50%",
               "none",
               "10%",
               "25%",
               "50%",
               "none")),
               group = unique(persistence.table3$group),
                      variable = "Raccoon")

mortality.labels.df <- data.frame(mortality.labels= c("none","+1%","","","",
                      "+2%","","",
                      "","+5%","","",
                      ""),
                      group = unique(persistence.table3$group),
                      variable = "Raccoon")

persistence.table3 <- left_join(persistence.table3,
                                mortality.labels.df,
                                by = c("group", "variable"))

persistence.table3 <- left_join(persistence.table3,
                                rehab.labels.df,
                                by = c("group", "variable"))

# Order high probability of persistence at top of heat map, low persistence at bottom right of figure
persistence.table3$group <- factor(persistence.table3$group,
                                   levels = c("+5%  none", "+5%  +10%", "+5%  +25%",  "+5%  +50%",
                                              "+2%  none", "+2%  +10%",  "+2%  +25%","+2%  +50%",
                                              "+1%  none", "+1%  +10%", "+1%  +25%",  "+1%  +50%",
                                              "base none"))

# change NA's to ""
persistence.table3$mortality.labels[is.na(persistence.table3$mortality.labels)] <- ""
persistence.table3$rehab.labels[is.na(persistence.table3$rehab.labels)] <- ""

# Re-order species
persistence.table3$variable <- factor(persistence.table3$variable,
                                   levels = c( "Raccoon",
                                               "Painted Turtle",
                                               "Snapping Turtle", 
                                               "Blanding's Turtle",
                                               "Little Brown Bat"))

# Add ranges to help with formatting
x.offset.rehab = 0.5
x.offset.mortality = 1.5
x.range =c(-0.5,5)
y.range = c(1,14)
y.label.height =  14

# Add exintction prob (1 - persistence prob)
persistence.table3$ext.risk <- 1-persistence.table3$value

# ggplot
heatmap.plot <- ggplot(persistence.table3, aes(x = variable, y = group)) + # plot variable by group
  # ggtitle("There goes my afternoon....\nbut at least it worked") +
  geom_tile(aes(fill = value)) + # background colours are mapped according to the value column
  geom_text(aes(fill = persistence.table3$value, label = round(persistence.table3$value, 2))) + # round to reduce long numbers
  scale_fill_gradient2(low = "red1",  mid = "white", high ="blue1", # scale from red (high extinction) to blue (low extinction)
                       midpoint = 0.5, 
                       space = "rgb",
                       guide = guide_colourbar(reverse = TRUE)) + 
  theme(panel.grid.major.x = element_blank(), # no gridlines
        panel.grid.minor.x = element_blank(), 
        panel.grid.major.y = element_blank(), 
        panel.grid.minor.y = element_blank(),
        panel.background = element_rect(fill="white"), 
        axis.text.x = element_text(angle = 25, size = 12, 
                                   face = "bold", colour = "black",
                                   hjust = 0),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.title=element_text(face = "bold", size = 14),
        legend.position = "bottom") + 
  geom_hline(yintercept  = 4.5, col = "black") +
  geom_hline(yintercept  = 8.5, col = "black") +
  geom_hline(yintercept  = 12.5, col = "black") +
  
  geom_vline(xintercept = -0.4, col = "black") +
  scale_x_discrete(name = "", position = "top")+
  # scale_y_discrete(name = "", labels = rehab.labels) +
  labs(fill="Probability of persistence") +
  geom_text(x = -1,# Set the position of the text to always be at '-0.5'
                hjust = 0,
            # size = 12,
                aes(label = mortality.labels)) +
   geom_text(x = 0,# Set the position of the text to always be at '-0.5'
                hjust = 0,
            # size = 12,
                aes(label = rehab.labels)) +
  annotate("text", label = "Additional\nmortality", x = -1, y = 14.25,
           colour = "black", face = "bold", hjust = 0.25) +
  annotate("text", label = "Rehabilitation", x = 0, y = 14,
           colour = "black", face = "bold", hjust = 0.25) +
  coord_cartesian(clip = 'off',
                  xlim = x.range, ylim = y.range) +
  theme(plot.margin = unit(c(1,3,1,3), "lines"))
heatmap.plot

# Save plot
ggsave(heatmap.plot, file = "Figures/Figure4_persistenceheatmap.png",
       height = 8, width = 8, dpi = 500)

```
